---
title: "Meta-analysis in R"
format: html
editor: visual
author: "Matthew Grainger"
---

## Introduction

This exercise introduces meta-analysis concepts and methods, focusing on how they apply to ecological data. You will work with R to calculate effect sizes, understand fixed and random effects models, explore multi-level meta-analysis, and visualise results.

### 1. Effect Sizes

#### Objective: Learn about effect sizes and how to compute them for ecological studies

[Effect sizes](https://matthewbjane.quarto.pub/Guide-to-Effect-Sizes-and-Confidence-Intervals.pdf "Detailed overview of different effect sizes") are key to synthesising study results in meta-analysis, as they provide a standardised way of comparing findings across studies. Common effect sizes include the standardised mean difference (Hedges' g in ecology), correlation coefficient, and log-transformed response ratio (lnRR).

Example Code:

We have data from six fictional studies that examined species abundance before and after restoration efforts. The dataset includes: - Sample sizes before and after restoration (`n1` and `n2`). - Mean species abundance before and after (`m1` and `m2`). - Standard deviations of abundance before and after (`sd1` and `sd2`).

| Study | Sample Size (Before) | Sample Size (After) | Mean Abundance (Before) | Mean Abundance (After) | SD (Before) | SD (After) |
|----|----|----|----|----|----|----|
| Smith_2015 | 30 | 28 | 5.2 | 3.9 | 1.1 | 0.9 |
| Johnson_2017 | 50 | 48 | 6.1 | 5.4 | 1.3 | 1.1 |
| Lee_2018 | 45 | 44 | 5.9 | 5.3 | 1.0 | 0.8 |
| Gomez_2016 | 32 | 30 | 4.8 | 4.2 | 1.2 | 1.0 |
| Patel_2019 | 40 | 39 | 5.5 | 4.9 | 1.4 | 1.2 |
| Chen_2020 | 35 | 33 | 5.1 | 4.5 | 1.2 | 1.1 |

The **standardised mean difference (SMD)** between restored and control sites is calculated as follows:

$$\text{SMD} = \frac{\text{mean}_{\text{restored}} - \text{mean}_{\text{control}}}{\text{pooled standard deviation}}$$ Where the **pooled standard deviation** is calculated as:

$$2SD_{\text{pooled}} = \sqrt{\frac{(n_{\text{restored}} - 1) \cdot SD_{\text{restored}}^2 + (n_{\text{control}} - 1) \cdot SD_{\text{control}}^2}{n_{\text{restored}} + n_{\text{control}} - 2}}$$

In ecology we often have small sample sizes, so we use Hedges g to "correct" for small samples. The R package `metafor` calculates this for us using the `escalc` function.

```{r}
#| echo: true
#| message: false
#| warning: false
#| label: get_data

# Install metafor if you need to
# install.packages("metafor")
# Load metafor
library(metafor)

# Example dataset of ecological studies with mean outcomes and standard deviations
data <- data.frame(
study = c("Smith_2015", "Johnson_2017", "Lee_2018", "Gomez_2016", "Patel_2019", "Chen_2020"),
n1 = c(30, 50, 45, 32, 40, 35),
n2 = c(28, 48, 44, 30, 39, 33),
m1 = c(5.2, 6.1, 5.9, 4.8, 5.5, 5.1),
m2 = c(3.9, 5.4, 5.3, 4.2, 4.9, 4.5),
sd1 = c(1.1, 1.3, 1.0, 1.2, 1.4, 1.2),
sd2 = c(0.9, 1.1, 0.8, 1.0, 1.2, 1.1)
)

# Calculate standardised mean differences
data$yi <- escalc(measure="SMD", m1i=m1, sd1i=sd1, n1i=n1, m2i=m2, sd2i=sd2, n2i=n2, data=data)$yi
data$vi <- escalc(measure="SMD", m1i=m1, sd1i=sd1, n1i=n1, m2i=m2, sd2i=sd2, n2i=n2, data=data)$vi

data
```

### 2. Fixed & Random Effects Meta-Analysis

#### Objective: Understand and compare fixed-effect and random-effects models

In a fixed-effect model, we assume that all studies estimate the same true effect. In contrast, a random-effects model assumes that each study estimates its own effect size, accounting for between-study variability.

Fit both models and compare results.

```{r}
#| echo: true
#| message: false
#| warning: false
#| label: compare_models

# Fixed-effect model
fixed_model <- rma(yi, vi, data = data, method="FE")
summary(fixed_model)
# Random-effects model
random_model <- rma(yi, vi, data = data, method="REML")
summary(random_model)

```

::: {.callout-note icon="false"}
## Discussion

Why might a random-effects model be more suitable for ecological data? What does the presence of between-study variance tell us?
:::

### 3.Multi-Level Meta-Analysis

#### Objective: Explore multi-level meta-analysis to handle nested data structures (e.g., multiple effect sizes per study).

Multi-level meta-analysis allows us to account for non-independence in data, such as when studies report multiple effect sizes, or when species are related to each other.

Here we generate a simulated dataset with nested effect sizes and use `metafor` to fit a multi-level model.

```{r}
# Simulated nested dataset
multi_data <- data.frame(
  study = rep(c("Smith_2015", "Johnson_2017", "Lee_2018", "Gomez_2016", "Patel_2019", "Chen_2020"), each = 2),
  effect = rnorm(12, 0.5, 0.2),
  vi = runif(12, 0.1, 0.2)
)

# Multi-level meta-analysis
multi_model <- rma.mv(effect, vi, random = ~ 1 | study, data = multi_data)
summary(multi_model)
```

::: {.callout-note icon="false"}
## Discussion:

Why might multi-level meta-analysis be relevant for ecology? How does this approach help with dependencies in data?
:::

### 4.Visualising Meta-Analysis Results

#### Objective: Learn to create orchard plots to visualise meta-analysis results

Orchard plots provide a clear visualisation of effect sizes and heterogeneity, making it easier to interpret meta-analysis outcomes.

Use `orchard_plot()` to visualise the results from the random-effects model and multi-level model.

```{r}
# install the package from GitHub
# We need the remotes package first to do this
# install.packages("remotes")
# remotes::install_github("daniel1noble/orchaRd")
# Load library orchaRd
library(orchaRd)
# Orchard plot for the random-effects model
orchard_plot(random_model, xlab = "Effect Size (Standardised Mean Difference)",group="study")

# Orchard plot for the multi-level model
orchard_plot(multi_model, xlab = "Effect Size (Standardised Mean Difference)",group="study")

```

::: {.callout-note icon="false"}
## Discussion:

How does the plot help in interpreting effect sizes? What additional insights do you get from visualisations compared to numerical summaries?

When might you choose a fixed-effect model over a random-effects model in ecology?

How can multi-level meta-analysis account for the complexities of ecological data?

What are the benefits and limitations of visualizations in meta-analysis?
:::

### 5. Cumulative meta-analysis

#### Objective: Learn to run a cumulative meta-analysis in R

A **cumulative meta-analysis** examines how the overall effect size estimate changes as each study is added sequentially, usually sorted by publication date. This approach can reveal trends over time, showing whether the effect size has stabilised as more studies are included.

Let's go through an example of performing a cumulative meta-analysis with the `metafor` package in R.

##### Step 1: Create the Dataset

First, we will create a dataset with a `year` variable to sort studies in chronological order for the cumulative analysis.

```{r}
#| echo: true
#| message: false
#| warning: false
data <- data.frame(
  study_id = 1:10,
  year = c(2000, 2002, 2004, 2005, 2006, 2008, 2010, 2012, 2015, 2018),  # Publication year
  yi = c(0.8, 0.7, 0.5, 0.3, 0.1, -0.1, -0.3, -0.5, -0.7, -0.9),  # Effect sizes (Hedges' g)
  vi = c(0.02, 0.03, 0.025, 0.04, 0.018, 0.027, 0.035, 0.033, 0.02, 0.03)  # Variances
)

# Sort data by year for cumulative meta-analysis
data <- data[order(data$year), ]

```

##### Step 2: Perform Cumulative Meta-Analysis

Now, we can use the `cumul()` function from `metafor` to perform a cumulative meta-analysis. This function recalculates the meta-analysis result each time a new study is added.

```{r}
#| echo: true
#| message: false
#| warning: false
# Fit an initial random-effects model 
random_effect_model <- rma(yi = yi, vi = vi, data = data, method = "REML")  # Perform cumulative meta-analysis 
cumulative_results <- cumul(random_effect_model, order = data$year)  # View cumulative meta-analysis 
cumulative_results

```

##### Step 3: Plot the Cumulative Meta-Analysis

You can visualize the cumulative effect size estimates over time using `forest()`.

```{r}
#| echo: true
#| message: false
#| warning: false
# Plot cumulative meta-analysis results 
forest(cumulative_results, xlab = "Cumulative Hedges' g", refline = 0, cex = 0.8)
```

##### Interpretation

As you add each study (from the earliest to the most recent), you will see how the cumulative estimate of Hedges' g changes. In ecology, this can highlight if early results were consistent with later studies or if new findings led to significant shifts in the overall effect estimate. This approach provides insight into whether further studies are likely to meaningfully alter the cumulative effect size or if it has stabilised.

::: {.callout-note icon="false"}
## Discussion:

Have a look at [this paper](https://onlinelibrary.wiley.com/doi/full/10.1002/jrsm.1691). What do the results of this study mean for evidence-informed decision making in ecology? What solutions can you think of to mitigate for this issue?
:::
